---
title: "Testing data points distribution"
# author: "R.Andres Castaneda"
execute:
  echo: false
  output: false
format: 
   html:
     df-print: paged
     fontsize: 1.1em
     linestretch: 1.7
     theme:
       light: flatly
       dark: darkly
editor: visual
---

## No monotonic distribution in welfare average

```{r}
#| label:  libraries

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
#                   Load Libraries   ---------
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

library(data.table)
library(here)
library(ggplot2)
source("R/init.R")


```

```{r}
#| label: set-up


#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
#Set up   ---------
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
wd <- 
  glue::glue("data/album/world_{nq}bin_{py}PPP.qs") |> 
  qs::qread()

setorderv(wd,
          c(
            "country_code",
            "year",
            "welfare_type",
            "reporting_level",
            "percentile"
          ))

vars  <- c("avg_welfare", "welfare_share", "percentile")
dvars <- paste0("d", vars)
tvars <- paste0("t", vars)

wd[,
  (dvars) := lapply(.SD, \(.x) {
                  y <- .x - shift(.x)
                  round(y, 5)
                }
                ),
  by = .(country_code,
        reporting_level,
        welfare_type,
        year),
  .SDcols = vars
  ][, 
    (tvars) := 
      lapply(.SD, \(.d) {
      .d < 0
        }
      ),
    by = .(country_code,
        reporting_level,
        welfare_type,
        year),
  .SDcols = dvars
    ]

```

```{r}
#| label: avg-welfare
aw <- 
  wd[tavg_welfare == TRUE, 
     .(bins = toString(percentile)), 
     by = c("country_code", "year")
     ]

pfw <- pipload::pip_load_aux("pfw")
# pfw[, year := as.character(year)]

dt <-
  merge(x = aw,
        y = pfw,
        by = c("country_code", "year"))

use_vars <- grep("^use_", names(dt), value = TRUE) |>
  c("country_code", "year", "bins")

dt[, data_type := fcase(use_imputed == 1, "imputed",
                        use_microdata == 1, "microdata", 
                        use_groupdata == 1, "groupdata", 
                        use_bin == 1, "bin", 
                        default = "")]
```

Data points for which welfare average in bin `b` is lower than that in `b-1`. Only `r dt[data_type == "microdata", .N]` cases come from microdata, and `r dt[data_type != "microdata", .N]` from bin data.

```{r}
#| output: true

dt[, c("country_code", "year", "bins", "data_type")]
```

```{r}

de <- 
  wd[twelfare_share == TRUE, 
    c("country_code", "year", "percentile")]

```

The main problem is due to too many repeated values in the misbehaved microdata. See, for example, all the jumps in the histogram of ARG 1994 in contrast to a better-behaved dataset like COL 2017.

```{r}
#| output: true
 
dt   <- pipload::pip_load_cache(c("COL", "ARG"),
                                c(2017, 1994),
                                verbose = FALSE)

dt <- dt[!(country_code == "ARG" & surveyid_year == 2017)]

setorder(dt, country_code, surveyid_year, imputation_id, reporting_level, welfare_ppp, weight)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
## bins ant totals --------

dt[,
   # get bins and total pop and welfare
   `:=`(
     bin = wbpip:::md_compute_bins(welfare_ppp,
                                   weight,
                                   nbins = nq,
                                   output = "simple"),
     tot_pop = sum(weight),
     tot_wlf = sum(welfare_ppp*weight)
   ),
   by = c("country_code", "surveyid_year", "imputation_id", "reporting_level")
]


ggplot(dt[bin < 60], aes(x=welfare_ppp)) +
  geom_histogram(aes(y=..density..), colour="black", fill="white")+
  geom_density(alpha=.2, fill="#FF6666") +
  facet_wrap(country_code + surveyid_year ~ ., 
             scales = "free")
```

When the same welfare value is repeated across bins, we may find that the threshold value (i.e., percentile) of bin `b` is the same as the welfare average of `b+1` and, sometimes, part of the values of `b+2`.

```{r}
#| label: same-welfare
#| output: true
#| tbl-cap-location: bottom
#| tbl-cap: "ARG 1994"


wd[country_code == "ARG" & 
     year == 1994 &
     between(percentile, 47, 49), 
   c("percentile", "avg_welfare", "percentile")
   ]


```

This problematic because we get inconsistencies like having the same headcount for different bins.

```{r}
#| output: true
popshares <- seq(from = .47, to = .5, by = .01)

hcs <- 
  sapply(popshares, \(.) {
    x <- pipr::get_stats("ARG", 1994, popshare = .)
    setDT(x)
    x[, headcount]
  })

data.table(popshare = popshares, 
           headcount = hcs)
```

## Non-monotonic distribution in welfare share

```{r}
#| label: shr-welfare
sw <- 
  wd[twelfare_share == TRUE, 
     .(bins = toString(percentile)), 
     by = c("country_code", "year")] 

df <-
  merge(x = sw,
        y = pfw,
        by = c("country_code", "year"))

use_vars <- grep("^use_", names(df), value = TRUE) |>
  c("country_code", "year", "bins")

df[, data_type := fcase(use_imputed == 1, "imputed",
                        use_microdata == 1, "microdata", 
                        use_groupdata == 1, "groupdata", 
                        use_bin == 1, "bin", 
                        default = "")]


df[, c("country_code", "year", "bins", "data_type")]
```

```{r}
#| label: pfw-treatment

fpf <- pfw[, .(country_code,
               surveyid_year,
               reporting_year,
               welfare_type
)]

fpf[,
    id := paste(country_code, 
                reporting_year, 
                welfare_type, 
                sep = "_")]
```

```{r}
#| label: arg-20-wlf-shr
cc <- "ARG"
yr <- 2002

fpf <- fpf[country_code == cc & 
             surveyid_year == yr]

pl <- fpf |>
  split(by = "id")

pl <- pl[[1]]

ar <- get_micro_dist(pl)
```

In most of the cases where the welfare share in `b` is smaller than in `b-1`, the differences is very small, mainly due to the way we create bins.

```{r}
#| label: plot-sh-wlf
#| output: true

we <- wd[dwelfare_share < 0]
we[, dws := -1*dwelfare_share]
sdws <- we[, sd(dws, na.rm = TRUE)]
mnws <- we[, mean(dws, na.rm = TRUE)]

ggplot( we[dws < (mnws + 2*sdws)], 
        aes(x = dwelfare_share)) +
  geom_histogram(aes(y  = ..density..), 
                 colour = "black", 
                 fill   = "white") +
  geom_density(alpha    = .2, 
               fill     = "#FF6666")



```

## Specific data

```{r}

dt   <- pipload::pip_load_cache(c("CIV"),
                                c(1998, 2002, 2008),
                                verbose = FALSE)

setorder(dt, country_code, surveyid_year, imputation_id, reporting_level, welfare_ppp, weight)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
## bins ant totals --------

dt[,
   # get bins and total pop and welfare
   `:=`(
     bin = wbpip:::md_compute_bins(welfare_ppp,
                                   weight,
                                   nbins = nq,
                                   output = "simple"),
     tot_pop = sum(weight),
     tot_wlf = sum(welfare_ppp*weight)
   ),
   by = c("country_code", "surveyid_year", "imputation_id", "reporting_level")
]


ggplot(dt[bin < 60], aes(x=welfare_ppp)) +
  geom_histogram(aes(y = after_stat(density)), colour="black", fill="white")+
  geom_density(alpha=.2, fill="#FF6666") +
  facet_wrap(country_code + surveyid_year ~ ., 
             scales = "free")


civ <- dt[ survey_year == 1998 & between(bin, 57,60), c("welfare_ppp", "weight") ]


ggplot(civ, aes(x = weight)) +
  geom_histogram(aes(y = after_stat(count)), colour = "black", fill="white")+
  geom_density(alpha=.2, fill="#FF6666")

  

```
